Letter: Performance analysis of \\ high-resolution ice sheet simulations

### 2022-06-25 paper/perfmod.tex
edits first two sections

\abstract{High spatial resolution numerical ice sheet models, with horizontal mesh/grid resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  These models usually require many time steps, usually shorter than climate-coupling time scales, because ice thickness is updated following each velocity solution.  That is, practical performance in the high-resolution limit is degraded by the stability conditions imposed by such explicit time-stepping.  In this short note, regarding the shallow ice approximation and Stokes models as stress balance end members, we attempt to clarify numerical model performance by quantifying the simulation cost per model year in terms of the mesh resolution and number of degrees of freedom.  The performance of coming generations of implicit time-stepping numerical models is then assessed.  Our main result, Table \ref{tab:performancemodel}, emphasizes the key role played by solver performance in Stokes models.}

@@

Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and quantification of long-term glacial erosion \citep{SeguinotDelaney2021}, among other applications.  It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows suggest the need for finer resolution.  Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used at ice sheet scale.

Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow and non-hydrostatic Stokes approximation.  With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the last-computed velocity field.  Therefore ice thickness and surface elevation, which are functionally-equivalent variables for this purpose, are updated after the velocity is fixed.  This geometry-update operation, using the mass continuity or surface kinematical equations \citep{GreveBlatter2009}, enables interaction between the ice sheet and the surrounding climate.  Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.

These standard schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  Now, at least for simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  The stability limitations of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening the steps \citep{LofgrenAhlkronaHelanow2021}.

However, actual \emph{implicit} time-stepping should also be considered.  Here the velocity and geometry are updated simultaneously by solving a coupled mass-momentum problem.  While implicit time-stepping always requires the solution of systems of equations at each step \citep{LeVeque2007}, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.  The problem is of free-boundary type, both in map-plane (horizontal) \citep{SchoofHewitt2013} and (easily-resolved) vertical directions.

A successful implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.  The same reference shows how the steady-state problem \citep[see][for example]{JouvetBueler2012} can often be solved; observe that steady-state solutions correspond to an implicit time step of infinite duration.  The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.  However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve for a free boundary illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.

This is the complicated context in which the current note relates time-stepping and stress-balance choices to computational effort.  While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected quantitatively to the extent possible.  A simplified performance (computational effort) analysis can expose the most important considerations and trade-offs.

@@

The reader is assumed to have access to the SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.  These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.

Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.  However, before presenting a performance analysis, the form of the mass continuity equation must be examined.  For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
\begin{equation}
\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a, \label{eq:masscontinuity}
\end{equation}
where $\bx=(x,y)$ denotes horizontal coordinates.

Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Indeed, ice sheet flow has no characteristic curves, as would a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.  Thus, as addressed by linearized analysis \citep{Chengetal2017,Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of advective and diffusive mechanisms.

Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation for bed elevation $b(\bx)$.  If a numerical thickness perturbation grows then the velocity $\bu$ often responds by increasing in a direction close to downhill ($-\gradx s$), a direction often correlated to $-\gradx H$.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, but the gravitational driving stress is always along $\gradx s$.  Equation \eqref{eq:masscontinuity} therefore has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$'', though a stress balance solution is required to evaluate this function.  A numerical instability occurs when the ice thickness under/over-shoots its correct (i.e.~continuum) value because the numerical velocity response is too strong.

\newcommand{\nn}{\text{n}}
The mass continuity equation is not wholely diffusive for Stokes dynamics, but the diffusive nature dominates in the small-parameter limit which generates the SIA.  In the isothermal case we see the following, revised mass continuity equation:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity and $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.  Typically $\nn\approx 3$.}

Equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics.  However, the same diffusivity $d$, an essentially geometric quantity, can be computed.  For grounded ice sheets one observes that large values of $d$ often indicate locations of unstable mode growth if explicit time-steps are chosen too large.

>>>>

\abstract{High spatial resolution numerical ice sheet models, with horizontal mesh/grid resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  These models usually require many time steps, shorter than climate-coupling time scales, because ice thickness is updated following each velocity solution.  High-resolution performance is degraded by the stability restrictions of such explicit time-stepping.  This short note, which considers the shallow ice approximation and Stokes models as stress-balance end members, attempts to clarify numerical model performance by quantifying simulation cost per model year in terms of mesh resolution and the number of degrees of freedom.  The prospective performance of implicit time-stepping numerical models is assessed.  The main result, Table \ref{tab:performancemodel}, emphasizes the key role played by the performance of stress-balance and implicit-step solvers.}

@@

Numerical ice sheet (glacier) models with evolving ice geometry are now in routine use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and evaluation of long-term glacial erosion rates \citep{SeguinotDelaney2021}, among other applications.  It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results, but narrow outlet glacier flows need finer resolution.  Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used at ice sheet scale.

Current-generation ice sheet models use a variety of stress balances, from the simplest shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow and non-hydrostatic Stokes approximation.  With very few exceptions, however, current models alternate between solving the stress balance for velocity, using the geometry determined by the previous time step, and then updating the geometry using the just-computed velocity field.  Therefore ice thickness and surface elevation, functionally-equivalent variables for this purpose, are updated after velocity is fixed.  Interaction between the ice sheet and the surrounding climate occurs during the geometry-update operation via the mass continuity or surface kinematical equations \citep{GreveBlatter2009}.  Such climatic coupling occurs through surface mass balance, sub-shelf (basal) mass balance, and calving processes in particular.

Thus these standard schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system which describes the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use the velocity computed from the geometry at the previous time step \citep[for example]{Chengetal2017}.}  At least for simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}, and stability conditions of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.

However, actual \emph{implicit} time-stepping \citep{Bueler2021,LeVeque2007} should also be considered.  Here the velocity and geometry are updated by simultaneously solving mass and momentum conservation equations.  While implicit time-stepping always requires the solution of systems of equations at each step \citep{LeVeque2007}, for ice sheets an implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.  The problem is of free-boundary type, both in map-plane (horizontal) \citep{SchoofHewitt2013} and (easily-resolved) vertical directions.

An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}, and the same reference shows how steady-state problem \citep[compare][]{JouvetBueler2012} can often be solved, demonstrating unconditional stability.  (Observe that steady-state solutions correspond to an implicit time step of infinite duration.)  The corresponding problem for the Stokes equations has not, to the author's knowledge, yet been attempted.  However, important early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} to solve for a free boundary illuminates some of the techniques and difficulties needed to make such a strategy work for a membrane-stress-resolving balance.

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  A simplified performance (computational work) analysis can expose the most important considerations and trade-offs.  While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping is applied, along with advanced solver techniques like multigrid \citep{Briggsetal2000}, this belief should be inspected quantitatively to the extent possible.

@@

The reader is assumed to have access to the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.  These (continuum) models are regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}.

Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.  However, before presenting a performance analysis, the form of the mass continuity equation must be examined.  For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged horizontal velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
\begin{equation}
\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a, \label{eq:masscontinuity}
\end{equation}
where $\bx=(x,y)$ denotes horizontal coordinates.

Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, especially regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Indeed, ice sheet flow has no characteristic curves, as would \eqref{eq:masscontinuity} if it were a true advection, because the velocity $\bu$ actually depends on the gradient of thickness through the stress balance.  Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.

Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  Under a numerical thickness perturbation the velocity $\bu$ often responds by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational driving stress source term is along $\gradx s$.  In fact equation \eqref{eq:masscontinuity} has velocity $\bu$ which is a non-local function $\bu(H,\grad_x s)$, though a stress balance solution is required to evaluate it.  A numerical instability occurs when the ice thickness under/over-shoots its correct value because the numerical velocity response from evaluating this non-local function is too strong.

\newcommand{\nn}{\text{n}}
Neither is the mass continuity equation wholely diffusive for Stokes dynamics.  A diffusive description is valid, however, in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}.  In the isothermal case we see a revised mass continuity equation:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity and $C$ is a positive constant.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ where $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.  Typically $\nn\approx 3$.}  While equation \eqref{eq:siamasscontinuity} does not hold for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed.  For grounded ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.
